Fabric-Elasticity Relationships in Osteogenesis Imperfecta (OI)

### 2021-05-18 09_Article/Simon_2021_Bone_Draft.tex
Add motion artefact grading and BMD histograms

\begin{abstract}
	Osteogenesis imperfecta (OI) is the most common inherited form of bone fragility. This disease, also called "brittle bone disease", is characterized by impaired synthesis of type I collagen leading to fragile bones. In this study, 50 OI diagnosed people were compared to 120 healthy controls using second generation high resolution peripheral computed tomography (HR-pQCT) scans of the distal tibia. Such scans are usually used in finite element analysis (FEA) for bone strength estimation and fracture risk assessment. The homogenized FE (hFE) scheme uses HR-pQCT bone volume fraction (BV/TV) and anisotropy information (fabric) and allows to assess bone strength within a reasonable computation time using fabric-elasticity relationships. Thus, the aim of this study is to investigate fabric-elasticity relationships in OI compared to healthy controls. For this, six regions of interest (ROIs) were selected per individual in a common anatomical region. A first age \& gender matching allowed to perform a morphometric analysis and compare the samples with literature. Then, stiffness tensors of ROIs was computed by FEA and multiple linear regression was performed on the Zysset-Curnier orthotropic model to compare groups using 5 parameters. This fit was first executed using the complete data sets separately. Then, data were filtered according to a fixed threshold for a defined coefficient of variation (CV) assessing the ROI heterogeneity. A second fit was performed on these filtered data sets and a third fit was done on BV/TV \& fabric anisotropy (DA) matched data sets. This last matched data set was grouped to compare with literature \si{\micro}CT scans. Morphometric analysis shown results in agreement with literature, namely significant differences between groups for BV/TV, trabecular number (Tb N), trabecular spacing (Tb Sp), trabecular spacing standard deviation (Tb Sp SD) but no significant differences for trabecular thickness (Tb Th). The multiple linear regressions shown that OI stiffness range reach lower values than healthy controls and the fit tends of overestimate the stiffness at lower values. Filtering out high heterogeneous ROIs removes low stiffness ROIs and lead to similar fit quality for both OI and healthy group. The BV/TV \& DA matched data sets shown no significant differences between OI and healthy individuals and stiffness constants are lower at 61 \si{\micro}m resolution HR-pQCT scans than for higher resolution 36 \si{\micro}m \si{\micro}CT scans. Our main conclusions are that lower scan resolution lead to lower stiffness because the structure obtained is not as optimized as in reality. Finally, if trabecular bone is homogeneous enough ROIs inside ROI, there no reason to suppose different fabric-elasticity relationships between OI and healthy individual.
\end{abstract}

@@

Osteogenesis imperfecta (OI) is the most common inherited form of bone fragility \cite{Tournis2018}. It was estimated to concern about 1/13'500 births, less severe forms being not accounted in this estimation as they are recognized later in life \cite{Lindahl2015}. Therefore, OI is considered as a rare metabolic bone disorder. Sillence classification \cite{Sillence1979} allows to define four main types of OI based on the clinical phenotype:
\begin{itemize}
	\item Type I: less severe
	\item Type II: lethal at most shortly after birth
	\item Type III: most severe surviving form
	\item Type IV: intermediate severity
\end{itemize}
Also known as "brittle bone disease", OI is characterized by impaired synthesis of type I collagen, leading to such brittle and fragile bones \cite{LIM2017}.\\

Bone fragility in OI is complex and not totally understood. High resolution peripheral quantitative tomography (HR-pQCT) scans allow to perform \textit{in vivo} assessment of trabecular architecture and volumetric BMD in distal radius and tibia \cite{Boutroy2005}. Moreover, the bony microstructure obtained from HR-pQCT can be used for finite element analysis (FEA) to predict mechanical properties \cite{Boutroy2008}. Homogenized FE (hFE) is a FEA scheme including volume fraction (BV/TV) and anisotropy information (fabric) of trabecular bone from HR-pQCT which to assess bone strength within reasonable computation time \cite{Pahr2009}. High correlations were found between patient-specific hFE and mechanical compression experiment of cadaveric samples \cite{Varga2011,AriasMoreno2019}. Thus, it could be legitimate to use hFE for OI patients bone strength estimation and fracture risk assessment with the assumption of identical fabric-elasticity relationships between OI and healthy individuals. Therefore, the present study aims to compare healthy with OI diagnosed individuals and to draw conclusions about fabric-elasticity relationships.

@@

\subsection{Subjects}
The healthy group include a total of 120 patients from a previous reproducibility study \cite{Schenk2020}. The sample is composed of 64 female and 56 male subjects aged between 20 and 92 years old with a median age of 26 [22 - 35] years. These subjects did not took any medication known to affect bone metabolism nor presented any prior osteoporosis fracture. The second group is composed of 35 female 15 male individuals leading to 50 OI diagnosed subjects. The youngest and oldest patients are 19 and 69 year old, respectively. The median age is 44 [33 - 55] years. 35 subjects were diagnosed with type I OI, 2 with type III, and 13 with type IV.

\subsection{HR-pQCT}
HR-pQCT scans (XtremeCTII, SCANCO Medical, Brütisellen,
Switzerland) were performed at the distal tibia on all patients from both groups. The OI subjects were scanned using the manufacturer's standard protocol. These scans were performed at the Shriners Hospital for Children and were shared to our group by the McGill University in Montreal. People of the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020}. The main differences with the manufacturer's standard protocol are the following:
\begin{enumerate}
	\item The reference line is positioned at the proximal margin of the dense structure formed by the tibia plafond instead of  the subchondral endplate of the ankle joint (standard clinical section) \cite{Whittier2020}.
	\item Three stacks were scanned proximal to the reference line instead of one stack at 22.5 mm proximal to the reference line for standard clinical section \cite{Whittier2020}.
\end{enumerate}

These differences are shown in Figure \ref{01_ClinicalSections}. For both group, each stack consist of 168 voxels and a resolution of 61 \si{\micro}m in the three principal directions which lead to a thickness about 10.2 mm for each stack. Scanning settings were a voltage of 60 kVp, 900 μA, 100 ms integration time for the healthy group as well as for the OI group.

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.225\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/01_OIClinicalSection}
		\caption{OI group}
		\label{01_OI}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.225\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/01_ControlClinicalSection}
		\caption{Healthy group}
		\label{01_Healthy}
	\end{subfigure}
	\caption{Clinical section scanned for both group}
	\label{01_ClinicalSections}
\end{figure}

\subsection{Image analysis}
The HR-pQCT scans were evaluated using the manufacturer's standard protocol. Then, the segmented images were used for further analysis.\\

@@

The ROI is defined as a cube of about 5.3 mm side length. This size was chosen to correspond to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and \citeauthor{Gross2013}\cite{Gross2013}. It was determined by \citeauthor{Daszkiewicz2017}\cite{Daszkiewicz2017} to be the optimal size to obtain accurate FEA results with minimal computation efforts.\\

The ROI analysis was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffstätten, Austria). The morphological parameters analyzed are: BV/TV, SMI, trabecular number (Tb. N.), trabecular thickness (Tb. Th.), trabecular spacing (Tb. Sp.), and the standard deviation of the trabecular spacing (Tb. Sp. SD). Moreover, ROI fabric was evaluated. The fabric tensor $\mathbf{M}$ is a positive-definite second-order tensor. It is build as shown in Equation \ref{Eq201} below:

\begin{equation}
	\mathbf{M} = \sum_{i=1}^{3}{m_i \mathbf{M}_i}
	\label{Eq201}
\end{equation}

where $m_i$ are the eigenvalues of $\mathbf{M}$ and $\mathbf{M}_i$ are the dyadic product of the corresponding eigenvectors $\mathbf{m}_i$ \cite{Cowin1985,Harrigan1985}. The fabric is then independent of BV/TV and normalized with $tr(\mathbf{M}) = 3$. To evaluate this tensor, the MIL method was used \cite{Moreno2014}. The fabric eigenvalues allow to compute the degree of anisotropy (DA) of the ROI.\\

A mechanical analysis was performed using \textsc{ABAQUS 6.14}. Each voxel of the segmented ROI was converted to mesh using fully integrated linear brick elements (C3D8) with a stiffness $E$ of 10 000 MPa and a Poisson's ratio $\nu$ of 0.3. The simulation consisted of 6 independent load cases, 3 uni-axial and 3 shear cases, using KUBCs. KUBCs were used according to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} were it was determined to be better than PMUBC to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model builds the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV $\rho$, fabric information $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$. The building of this tensor is shown in Equation \ref{Eq202}.

\begin{equation}
	\begin{split}
		\mathbb{S}(\rho,\mathbf{M}) & = \sum_{i=1}^{3} (\lambda_0 + 2\mu_0)\rho^k m_i^{2l} \mathbf{M}_i \otimes \mathbf{M}_i \\ & + \sum_{\substack{i,j=1\\i \neq j}}^{3} \lambda_0' \rho^k m_i^{l} m_j^{l} \mathbf{M}_i \otimes \mathbf{M}_j \\ & + \sum_{\substack{i,j=1\\i \neq j}}^{3} \mu_0 \rho^k m_i^{l} m_j^{l} \mathbf{M}_i \overline{\underline{\otimes}} \mathbf{M}_j
	\end{split}
	\label{Eq202}
\end{equation}

Where $\otimes$ and $\overline{\underline{\otimes}}$ are the dyadic and symmetric product of second order tensors, respectively. The Zysset-Curnier model is built with the assumption of orthotropy. However, the trabecular structure is not perfectly orthotropic. In order to assess the ROI heterogeneity a so-called coefficient of variation (CV) is computed as presented in \cite{Panyasantisuk2015}: the ROI is divided into eight identical subcubes, BV/TV is computed for each subcube and the CV is defined as the ratio between the standard deviation of these BV/TV and the mean value, see Equation \ref{Eq203} below:

@@

\subsection{Statistics}
The height morphological parameters analyzed (BV/TV, Tb. N., Tb. Th., Tb. Sp., and Tb. Sp. SD, SMI, DA, and CV) were compared for both group. As the initial groups do not have similar distributions in quality of age and sex, a matching was performed leading to identical mean and median age as well as identical gender distribution. Then, the selection of statistical test to perform was executed as follow:
\begin{enumerate}
	\item For each parameter, the median value between the six ROI of the same individual was computed. The median was preferred over the mean because it is less influenced by outliers.
	\item Normality of the distribution was assessed with QQ plot and Shapiro-Wilk test.
	\item If normality assumption was met, Bartlett test for equal variances was performed. Otherwise, data were log transformed to try to achieve normal distribution. If even after log transformation, normality assumption was not met, original data space was kept and Brown-Forsythe test was applied to assess the equal variance assumption.
	\item According to the previous results, t-test was performed if normal distribution and equal variance were met. If only the equal variances assumption was met, Mann-Whitney test was preferred. Finally, if none of these conditions could be assumed, a non-parametric permutation test was performed.
\end{enumerate}
The general significance level was set to 95\% for all tests. Confidence intervals in means difference was computed for t-tested variables to confirm p values. As Mann-Whitney test is performed on the median, only corresponding p value is given. Finally, non-parametric permutation test is less powerful but give an empirical 95\% exclusion range and a p value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

\subsection{Fit to model}
The stiffness tensors obtained from the mechanical simulations were transformed into fabric coordinate system and projected onto orthotropy, leading to 12 values. The resulting orthotropic stiffness tensors were then used to perform a multiple linear regression on the Zysset-Curnier model. Standard linear models assume independent and identically distributed (iid) variables. As this assumption is violated by the fact that six ROIs are analyzed by individual, a linear mixed-effect model was preferred. This last model, shown in Equation \ref{Eq204} in Laird-Ware form \cite{Laird1982}, takes into account the non-independence of ROIs from the same individual.

@@

The linear regression was performed on both group (healthy and OI) separately. To improve the fit quality, the data sets were filtered. The aim here is to filter out ROIs whose are too far from the assumption of orthotropy. Therefore, analogously to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}, a fixed threshold for the CV was used. To simplify comparison, the same value of 0.263 was fixed as exclusion criterion. Besides, the relation between BV/TV and CV was assessed using Spearman's correlation coefficient. Furthermore, to compare the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) between the groups, regression must be performed on identical value ranges. To do this, a matching was performed for BV/TV and DA to find corresponding control ROI for each OI in the filtered groups. Best correspondences were kept and duplicates were dropped. Finally, as the regression is performed in the log space, it is necessary to use identical exponent ($k$ and $l$) for both group, weighting identically BV/TV and DA between regressions, to compare stiffness constants. The exponents were determined by grouping healthy and OI for regression. Then a modified system is used to perform the fit on separated groups.\\

@@

\begin{sidewaystable*}
	\centering
	\caption{Summary of morphological analysis and comparison with literature. Values are presented as mean $\pm$ standard deviation when statistical test is performed on the means or median (inter-quartile range) when test is on medians. The study of \cite{Kocijan2015} presents n.s. for non-significant p value test result.}
	\label{Table2}
	\begin{tabular}{cccccccccc}
		\toprule
		\multirow{2}{*}{Variable} & \multirow{2}{*}{Group} & \multicolumn{2}{c}{Present study} & \multicolumn{2}{c}{\citeauthor{Folkestad2012}\cite{Folkestad2012}} & \multicolumn{2}{c}{\citeauthor{Kocijan2015}\cite{Kocijan2015}} & \multicolumn{2}{c}{\citeauthor{Rolvien2018}\cite{Rolvien2018}} \\
		& & Values & p value & Values & p value & Values & p value & Values & p value \\
		\midrule

@@

\subsection{Linear Regression}
Figure \ref{02_GeneralRegression} shows the result of regression for complete data sets of each group separately. The x axis represent the values of observed tensors from mechanical simulations. The y axis is the predicted values using the Zysset-Curnier model and the parameters obtained after performing the regression with linear mixed-effect model. The fitted line is represented by the dashed line. $\lambda_{ii}$ stands for the diagonal terms of normal components of $\mathbb{S}$ in Mandel notation\cite{MANDEL1965}, $\lambda_{ij}$ for the off-diagonal terms of normal components, and $\mu_{ij}$ for the shear components. For the healthy group, see Figure \ref{02_Healthy}, the fit is performed on 720 ROIs leading to 8640 points. The $R^2_{adj}$ is slightly above 0.95 and the NE is of 18\% $\pm$ 10\%. The regression of the OI group could only be performed on 294 ROIs as it was not possible to find suitable ROIs for one patient. Nevertheless, it leads to 3528 points, a $R^2_{adj}$ close to 0.85 and a NE of 62\% $\pm$ 233\%. It can be noticed that, as the values of observed tensor decreases, points tends to be further apart of the diagonal (dashed line). Moreover, the points coming from the tensors with lowest values are only above the diagonal. One last observation which should be noted is the range covered by the observed tensors. The range of the OI group is wider than the healthy group and goes further in low observed values.\\

@@

The CV in relation to BV/TV is shown in Figure \ref{02_CV_BVTV}. The OI data reach higher values of CV and stay at lower values of BV/TV than healthy data. Generally, the CV tends to increase with decreasing BV/TV. The Spearman coefficient is shown above the plot as value [95\% CI]. Its value is negative and strictly different from zero. Finally, the CV threshold value used to filter the data is represented by the dashed line. It can be observed that a relatively important part of OI data will be filtered out. On the other hand, relatively few healthy data will be removed by the filtering. Examples of extreme ROIs in terms of CV and BV/TV are available in Appendix \ref{A2}.\\

@@

		Present study & 166 & 0.21 $\pm$ 0.08 & 2.03 $\pm$ 0.21 & 2817 & 1640 & 2154 & 1.60 & 0.99 & 0.826 & 22 $\pm$ 12\\
		\bottomrule
	\end{tabular}
\end{table*}

>>>>

\begin{abstract}
	Osteogenesis Imperfecta (OI) is an inherited form of bone fragility. This disease, also called "brittle bone disease", is characterised by impaired synthesis of type I collagen, altered trabecular bone architecture and reduced bone mass that lead to fragile bones. High resolution peripheral computed tomography (HR-pQCT) is a powerful method to investigate bone morphology in the weight-bearing distal tibia and the 3D reconstructions can be exploited in either micro-finite element ($\mu$FE) or homogenised finite element (hFE) analysis for bone strength estimation. The hFE scheme uses local bone volume fraction (BV/TV) and anisotropy information (fabric) to compute healthy bone strength within a reasonable computation time using fabric-elasticity relationships. Thus, the aim of this study is to investigate fabric-elasticity relationships in OI trabecular bone compared to healthy controls.

	In this study, the morphology of distal tibias of 50 OI diagnosed people were compared to 120 healthy controls using second generation HR-pQCT. 
	Six regions of interest (ROIs) were selected per individual in a common anatomical region. A first age \& gender matching allowed to perform a morphometric analysis and compare the outcome with literature. Then, stiffness tensors of ROIs were computed by FEA and multiple linear regressions were performed on the Zysset-Curnier orthotropic model to compare the two groups using 5 parameters. After initial fits with all the samples of each group, the data were filtered according to a fixed threshold for a defined coefficient of variation (CV) assessing the ROI heterogeneity. Second fits were performed on these filtered data sets and then additional fits were done on BV/TV \& fabric anisotropy (DA) matched data to detect statistical differences between the two groups. This full and filtered data were in turn compared with previous results from $\mu$CT reconstructions obtained in other anatomical locations.

	In agreement with available literature, significantly lower BV/TV, trabecular number (Tb N), significantly higher trabecular spacing (Tb Sp), trabecular spacing standard deviation (Tb Sp SD), but no differences in trabecular thickness (Tb Th) were found between OI and controls. The stiffness of ROIs from OI bone reached lower values and the multilinear fabric-elasticity fits tended to overestimate the stiffness in the lower range. Filtering out highly heterogeneous ROIs removed these low stiffness ROIs and lead to similar correlation coefficients for both OI and healthy groups. Finally, the BV/TV \& DA matched data revealed no significant differences in fabric-elasticity parameters between OI and healthy individuals. Comparing with previous studies, the stiffness constants from 61 $\mu$m resolution HR-pQCT ROIs were lower than for 36 $\mu$m resolution $\mu$m CT images.

	In conclusion, despite the reduced regression parameters found for HR-pQCT images, the fabric-elasticity relationships between OI and healthy individuals are similar when the trabecular bone ROIs are sufficiently homogeneous to perform the mechanical analysis. Since highly heterogenous ROIs coincide with very low BV/TV, we expect them to play a minor role in hFE analysis of distal bone sections or parts.
\end{abstract}

@@

Osteogenesis imperfecta (OI) is an inherited form of bone fragility \cite{Tournis2018}. It was estimated to concern about 1/13'500 births, less severe forms being not accounted in this estimation as they are recognized later in life \cite{Lindahl2015}. Therefore, OI is considered as a rare metabolic bone disorder. Sillence classification \cite{Sillence1979} allows to define four main types of OI based on the clinical phenotype:
\begin{itemize}
	\item Type I: less severe
	\item Type II: lethal at most shortly after birth
	\item Type III: most severe surviving form
	\item Type IV: intermediate severity
\end{itemize}
Also known as "brittle bone disease", OI is characterized by impaired synthesis of type I collagen, leading to such brittle and fragile bones \cite{LIM2017}.\\

Bone fragility in OI is complex and not totally understood despite the investigations at different hierarchical levels. Multiple studies show that DXA areal BMD (aBMD) tends to be lower in OI than compared to healthy individuals \cite{Folkestad2012,Lindahl2015,Scheres2018}. The microstructure is different as well. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} shown that BV/TV and trabecular number in OI is lower than for healthy controls. Trabecular spacing and inhomogeneity are higher for OI diagnosed people but the trabecular thickness is not significantly different. At the ECM level, a recent study shown that OI bone tends to present higher modulus, ultimate stress and post-yield behavior than healthy bone in compression \cite{Indermaur2021}.\\

High resolution peripheral quantitative tomography (HR-pQCT) scans allow to perform \textit{in vivo} assessment of trabecular architecture and volumetric BMD in distal radius and tibia \cite{Boutroy2005}. Moreover, the bony microstructure obtained from HR-pQCT can be used for finite element analysis (FEA) to predict mechanical properties \cite{Boutroy2008}. Homogenized FE (hFE) is a FEA scheme including volume fraction (BV/TV) and anisotropy information (fabric) of trabecular bone from HR-pQCT which to assess bone strength within reasonable computation time \cite{Pahr2009}. High correlations were found between patient-specific hFE and mechanical compression experiment of cadaveric samples \cite{Varga2011,AriasMoreno2019}. Thus, it could be legitimate to use hFE for OI patients bone strength estimation and fracture risk assessment. However, HR-pQCT-based FEA rely on fabric-elasticity relationships. Therefore, the present study aims to compare trabecular bone microstructure of healthy and OI diagnosed individuals and to investigate the hypothesis of similar fabric-elasticity relationships.

@@

\subsection{Subjects}
The healthy group include a total of 120 patients from a previous reproducibility study performed at the University Department of osteoporosis in Bern \cite{Schenk2020}. The sample is composed of 64 female and 56 male subjects aged between 20 and 92 years old with a median age of 26 [22 - 35] years. These subjects did not took any medication known to affect bone metabolism nor presented any prior osteoporosis fracture. The second group was scanned at the Shriners Hospital for Children and images were shared to our group by the McGill University in Montreal. This group is composed of 35 female 15 male individuals leading to 50 OI diagnosed subjects. The youngest and oldest patients are 19 and 69 year old, respectively. The median age is 44 [33 - 55] years. 35 subjects were diagnosed with type I OI, 2 with type III, and 13 with type IV.

\subsection{HR-pQCT}
HR-pQCT scans (XtremeCTII, SCANCO Medical, Brütisellen,
Switzerland) were performed at the distal tibia on all patients from both groups. People of the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020} whereas OI subjects were scanned using the manufacturer's standard protocol. The main differences of the in-house protocol with the manufacturer's standard protocol are the following:
\begin{enumerate}
	\item The reference line is positioned at the proximal margin of the dense structure formed by the tibia plafond instead of  the subchondral endplate of the ankle joint (standard clinical section) \cite{Whittier2020}.
	\item Three stacks were scanned proximal to the reference line instead of one stack at 22.5 mm proximal to the reference line for standard clinical section \cite{Whittier2020}.
\end{enumerate}

These differences are shown in Figure \ref{01_ClinicalSections}. For both group, each stack consist of 168 voxels and a resolution of 61 \si{\micro}m in the three principal directions which lead to a thickness about 10.2 mm for each stack. Scanning settings were a voltage of 60 kVp, 900 μA, 100 ms integration time for the healthy group as well as for the OI group. For the healthy group, motion artefacts of first, middle and last slice were graded. The scale used start from 1 (no motion artefacts) to 5 (extreme motion artefacts) as proposed by the manufacturer \cite{Pialat2012}. The final grade of each scan was defined as the highest slice grade. For the OI group, as the scan consist of one stack, only one grade is attributed using the same scale as for the control group. Scans were then processed independently of their quality grading. A summary of the scans grading densities is shown in Figure \ref{01_MotionArtefacts}.

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.225\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/01_ControlClinicalSection}
		\caption{Healthy group}
		\label{01_Healthy}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.225\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/01_OIClinicalSection}
		\caption{OI group}
		\label{01_OI}
	\end{subfigure}
	\caption{\centering Clinical section scanned for both group}
	\label{01_ClinicalSections}
\end{figure}

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]
	{Pictures/01_MotionArtefacts}
	\caption{Summary of the motion artefacts grading. Histograms show density of each grade within both group.}
	\label{01_MotionArtefacts}
\end{figure}

\subsection{Image analysis}
The HR-pQCT scans were evaluated using the manufacturer's standard protocol. Then, mask segmented images were used for further analysis.\\

@@

The ROI is defined as a cube of 5.3 mm side length. This size was chosen to correspond to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and \citeauthor{Gross2013}\cite{Gross2013} which performed similar analysis with femur \si{\micro}CT scans. It was determined by \citeauthor{Zysset1998}\cite{Zysset1998} and \citeauthor{Daszkiewicz2017}\cite{Daszkiewicz2017} to be the optimal size to obtain accurate FEA results.\\

The morphological analysis of ROIs was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffstätten, Austria). The morphological parameters analyzed are: BV/TV, SMI, trabecular number (Tb. N.), trabecular thickness (Tb. Th.), trabecular spacing (Tb. Sp.), and the standard deviation of the trabecular spacing (Tb. Sp. SD). Moreover, ROI fabric was evaluated using MIL method \cite{Moreno2014}. The fabric tensor $\mathbf{M}$ is a positive-definite second-order tensor. It is build as shown in Equation \ref{Eq201} below:

\begin{equation}
	\mathbf{M} = \sum_{i=1}^{3}{m_i \mathbf{M}_i} = \sum_{i=1}^{3}{m_i \mathbf{m}_i \otimes \mathbf{m}_i}
	\label{Eq201}
\end{equation}

where $m_i$ are the eigenvalues of $\mathbf{M}$ and $\mathbf{M}_i$ are the dyadic product of the corresponding eigenvectors $\mathbf{m}_i$ \cite{Cowin1985,Harrigan1985}. The fabric is then independent of BV/TV and normalized with $tr(\mathbf{M}) = 3$. The fabric eigenvalues allow to compute the degree of anisotropy (DA) of the ROI by dividing the highest eigenvalue by the lowest one.\\

After ROI cleaning, i.e. deletion of unconnected region of bone material, an homogenized mechanical analysis was performed using \textsc{ABAQUS 6.14}. Each voxel of the cleaned ROI was converted to mesh using fully integrated linear brick elements (C3D8) with a stiffness $E$ of 10 000 MPa and a Poisson's ratio $\nu$ of 0.3. The simulation consisted of 6 independent load cases, 3 uni-axial and 3 shear cases, using KUBCs. KUBCs were used according to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}. Unlike PMUBCs, KUBCs do not require to rotate the ROI into fabric coordinate system which decrease the image quality. This homogenization process allows to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model builds the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV or $\rho$, fabric information $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$. The building of this tensor is shown in Equation \ref{Eq202}.\\

\begin{equation}
	\begin{split}
		&\mathbb{S}(\rho,\mathbf{M}) &=& \quad\sum_{i=1}^{3} \lambda_{ii} \mathbf{M}_i \otimes \mathbf{M}_i \\ &&&+ \sum_{\substack{i,j=1\\i \neq j}}^{3} \lambda_{ij} \mathbf{M}_i \otimes \mathbf{M}_j \\ &&&+ \sum_{\substack{i,j=1\\i \neq j}}^{3} \mu_{ij} \mathbf{M}_i \overline{\underline{\otimes}} \mathbf{M}_j \\
		&\text{With} &\\
		&\qquad\lambda_{ii} &=& \quad(\lambda_0 + 2\mu_0)\rho^k m_i^{2l} \\
		&\qquad\lambda_{ij} &=& \quad\lambda_0' \rho^k m_i^{l} m_j^{l} \\
		&\qquad\mu_{ij} &=& \quad\mu_0 \rho^k m_i^{l} m_j^{l} \\
	\end{split}
	\label{Eq202}
\end{equation}

Where $\otimes$ and $\overline{\underline{\otimes}}$ are the dyadic and symmetric product of second order tensors, respectively. The Zysset-Curnier model is built with the assumption of orthotropy and homogeneity. However, the trabecular structure is not perfectly homogeneous. In order to assess the ROI heterogeneity a so-called coefficient of variation (CV) is computed as presented in \cite{Panyasantisuk2015}: the ROI is divided into eight identical subcubes, BV/TV is computed for each subcube and the CV is defined as the ratio between the standard deviation of these BV/TV and the mean value, see Equation \ref{Eq203} below:

@@

\subsection{Statistics}
The morphological parameters analyzed (BV/TV, Tb. N., Tb. Th., Tb. Sp., and Tb. Sp. SD, SMI, DA, and CV) were compared for both group. As the initial groups do not have similar distributions of age and sex, a matching was performed leading to identical mean and median age as well as identical gender distribution. Then, the selection of statistical test to perform was executed as follow:
\begin{enumerate}
	\item For each parameter, the median value between the six ROI of the same individual was computed. The median was preferred over the mean because it is less influenced by outliers.
	\item Normality of the distribution was assessed with QQ plot and Shapiro-Wilk test.
	\item If normality assumption was met, Bartlett test for equal variances was performed. Otherwise, data were log transformed to try to achieve normal distribution. If even after log transformation, normality assumption was not met, original data space was kept and Brown-Forsythe test was applied to assess the equal variance assumption.
	\item According to the previous results, t-test was performed if normal distribution and equal variance were met. If only the equal variances assumption was met, Mann-Whitney test was preferred. Finally, if none of these conditions could be assumed, a non-parametric permutation test was performed.
\end{enumerate}
The general significance level was set to 95\% for all tests. Confidence intervals in means difference was computed for t-tested variables to confirm p values. As Mann-Whitney test is performed on the median, only corresponding p value is given. Finally, non-parametric permutation test is less powerful but give an empirical 95\% exclusion range and a p value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

\subsection{Fit to model}
The stiffness tensors obtained from the mechanical simulations were transformed into fabric coordinate system and projected onto orthotropy, leading to 12 components. The resulting orthotropic stiffness tensors were then used to perform a multiple linear regression on the Zysset-Curnier model. Standard linear models assume independent and identically distributed (iid) variables. As this assumption is violated by the fact that six ROIs are analyzed by individual, a linear mixed-effect model was preferred. This last model, shown in Equation \ref{Eq204} in Laird-Ware form \cite{Laird1982}, takes into account the non-independence of ROIs from the same individual. Detailed form of this model is presented in Appendix \ref{A1}.

@@

The linear regression was performed on both group (healthy and OI) separately. To improve the fit quality, the data sets were filtered. The aim here is to filter out ROIs whose are too far from the assumption of homogeneity. Therefore, analogously to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}, a fixed threshold for the CV was used. To simplify comparison, the same value of 0.263 was fixed as exclusion criterion. Besides, the relation between BV/TV and CV was assessed using Spearman's correlation coefficient. Furthermore, to compare the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) between the groups, regression must be performed on identical value ranges. To do this, a matching was performed for BV/TV and DA to find corresponding control ROI for each OI in the filtered groups. Best correspondences were kept and duplicates were dropped. Finally, as the regression is performed in the log space, it is necessary to use identical exponent ($k$ and $l$) for both group, weighting identically BV/TV and DA between regressions, to compare stiffness constants. The exponents were determined by grouping healthy and OI for regression. Then a modified system is used to perform the fit on separated groups.\\

@@

\begin{sidewaystable*}
	\centering
	\caption{Summary of the tibia ROIs morphological analysis and comparison with literature. Values are presented as mean $\pm$ standard deviation when statistical test is performed on the means or median (inter-quartile range) when test is on medians. The study of \cite{Kocijan2015} presents n.s. for non-significant p value test result.}
	\label{Table2}
	\begin{tabular}{cccccccccc}
		\toprule
		\multirow{2}{*}{Variable} & \multirow{2}{*}{Group} & \multicolumn{2}{c}{Present study} & \multicolumn{2}{c}{\citeauthor{Folkestad2012}\cite{Folkestad2012}} & \multicolumn{2}{c}{\citeauthor{Kocijan2015}\cite{Kocijan2015}} & \multicolumn{2}{c}{\citeauthor{Rolvien2018}\cite{Rolvien2018}} \\
		& & Values & p value & Values & p value & Values & p value & Values & p value \\
		\midrule

@@

\subsection{Linear Regressions}
Figure \ref{02_GeneralRegression} shows the result of regression for complete data sets of each group separately. The x axis represent the values of observed tensors from mechanical simulations. The y axis is the predicted values using the Zysset-Curnier model and the parameters obtained after performing the regression with linear mixed-effect model. The fitted line is represented by the dashed line. $\lambda_{ii}$ stands for the diagonal terms of normal components of $\mathbb{S}$ in Mandel notation\cite{MANDEL1965}, $\lambda_{ij}$ for the off-diagonal terms of normal components, and $\mu_{ij}$ for the shear components. For the healthy group, see Figure \ref{02_Healthy}, the fit is performed on 720 ROIs leading to 8640 points. The $R^2_{adj}$ is slightly above 0.95 and the NE is of 18\% $\pm$ 10\%. The regression of the OI group could only be performed on 294 ROIs as it was not possible to find suitable ROIs for one patient. Nevertheless, it leads to 3528 points, a $R^2_{adj}$ close to 0.85 and a NE of 62\% $\pm$ 233\%. It can be noticed that, as the values of observed tensor decreases, points tends to be further apart of the diagonal (dashed line). Moreover, the points coming from the tensors with lowest values are only above the diagonal. One last observation which should be noted is the range covered by the observed tensors. The range of the OI group is wider than the healthy group and ROIs with lower BV/TV present further low stiffness values.\\

@@

The CV in relation to BV/TV is shown in Figure \ref{02_CV_BVTV}. The OI data reach higher values of CV and stay at lower values of BV/TV than healthy data. Generally, the CV tends to increase with decreasing BV/TV. The Spearman coefficient is shown above the plot as value [95\% CI]. Its value is negative and strictly different from zero. Finally, the CV threshold value used to filter the data is represented by the dashed line. It can be observed that a relatively important part of OI data will be filtered out. On the other hand, relatively few healthy data will be removed by the filtering. Examples of extreme ROIs in terms of CV and BV/TV are shown in Appendix \ref{A2}.\\

@@

		\multirow{2}{*}{Present study} & 1014 & 0.23 $\pm$ 0.11 & 1.94 $\pm$ 0.26 & 2727 & 1655 & 2179 & 1.60 & 0.99 & 0.608 & 41 $\pm$ 182\\
		& 1014 & 0.23 $\pm$ 0.11 & 1.94 $\pm$ 0.26 & 5323 & 3232 & 4254 & 2.02 & 0.91 & 0.915 & 31 $\pm$ 105\\
		\bottomrule
	\end{tabular}
\end{table*}
